---
title: "Hurricane Analysis"
author: "Tyler Pollard"
date: "2024-08-25"
header-includes:
  - \usepackage{mathtools}
output:  
  github_document:
    html_preview: true
    includes: 
      in_header: _includes/head.html
    toc: true
    toc_depth: 3
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load libraries}
# Load Libraries ----
library(knitr)
library(data.table)
library(MASS)
library(tictoc)

# Bayesian
library(rjags)
library(caret)
library(bayesplot)
library(BayesFactor)
library(brms)
library(rstanarm)

# Data transforming and data viz
library(stringr)
library(plyr)
library(gt)
library(gtsummary)
library(tidyverse)
```

```{r Read in Data}
# Read in data ----
## Clean data ----
Stormdata <- fread("_data/E2_data.csv")
Stormdata$basin <- ifelse(Stormdata$basin == "atlantic", 1, 0)

## Create training and test data sets ----
training_data <- Stormdata |> filter(complete.cases(VMAX))
test_data <- Stormdata |> filter(!complete.cases(VMAX))
```

# Introduction

The data for this analysis are from [A Feed Forward Neural Network Based on Model Output Statistics for Short-Term Hurricane Intensity Prediction](https://journals.ametsoc.org/view/journals/wefo/34/4/waf-d-18-0173_1.xml). The data variables descriptions are available [here](https://github.com/TylerPollard410/Hurricane-Analysis/blob/main/docs/data_description.pdf). This paper uses deep learning to improve 24-hour ahead forecasts of hurricane intensity (maximum wind velocity, VMAX). The main prediction model is [HWRF](https://www.aoml.noaa.gov/hurricane-weather-research-forecast-model/), which is a mathematical model based on differential equations. In addition to the forecast, HWRF has many other state variables such as sea surface temperature, longitude, time of year, etc, that are usually discarded. In this analysis, we will determine if including these state variables can improve the HWRF prediction.

# Data Exploration (Shiny App)

To begin this analysis, I created a [Hurricane Analysis app](https://tylerpollard410.shinyapps.io/Hurricane_EDA/) to better understand and explore the data. The distribution of VMAX was plotted using the "Histogram" plot type, which revealed positive right skewness. The importance of spatial predictors like basin, Land, LAT, and LON can be viewed using the "Map" plot type. VMAX was plotted against all potential covariates and their transformed values to explore their relationship with the response and identify the best transformations to apply prior to model fitting.

# Model Description

The response variable, $Y_i$, is the observed VMAX of observation $i$ with StormID index $S_i$ of a given hurricane. The $p = 20$ covariates $X_{ij}$ that were fit to the model were LAT, LON, basin, Land, MINSLP, SHR_MAG, STM_SPD, SST, RHL0, CAPE1, CAPE3, SHTFL2, TCOND7002, INST2, CP1, TCONDSYM2, COUPLSYM3, HWFI, VMAX_OP_T0, and HWRF. All covariates were normalized using a Yeo-Johnson transformation with centering and scaling. Due to the positive right skewness of VMAX, models were compared for Log-normal and Gamma likelihoods both with and without random effects.

## Lognormal

$$
\begin{aligned}
Y_{i} &\sim \text{LogNormal}(\mu_i, \sigma^2) \\
\mu_i &= \beta_{0} + \sum_{j=1}^{p}X_{ij}\beta_{j} + \theta_{S_i} \\
\theta_{S_i} &\sim \text{Normal}(0, \tau^2)
\end{aligned}
$$

## Gamma

$$
\begin{aligned}
Y_{i} &\sim \text{Gamma}(\alpha, \alpha/\mu_i) \\
log(\mu_i) &= \beta_{0} + \sum_{j=1}^{p}X_{ij}\beta_{j} + \theta_{S_i} \\
\theta_{S_i} &\sim \text{Normal}(0, \tau^2)
\end{aligned}
$$

where $\beta_{j}$ is the effect of covariate $j$ with weakly informative priors $\beta_j \sim \text{Normal}(0,5)$. We set a non-informative prior on $\alpha \sim \text{InvGamma}(0.1, 0.1)$. In the random effects models, we set non-informative priors $\tau \sim \text{InvGamma}(0.1, 0.1)$.

# Model Comparisons

All four models above were fit to the data and were first checked that the models converged to an acceptable level. Once convergence was confirmed, the models were compared based on WAIC to determine the model with the best expected predictive accuracy. Based on Table 1, I chose to go forward with the slopes as random effects model because it had the smallest mean deviance which is a good sign that it will predict the best, although it was not as simple as the first model with constant slopes.

```{r WAIC}
load(file = "_data/waicList.RData")

waicListDf <- do.call(rbind, waicList)

```

## Model Selection Criteria

# Goodness of Fit

To indicate the goodness of fit of the model, posterior predictive checks were simulated as the data was being simulated. For every iteration of the MCMC, a draw of size $n$ was made from the posterior predictive distribution of $Y_i$ and the mean, standard deviation, and max were calculated to also get a draw from their distribution to compare directly back to the observed values of VMAX. The Bayesian p-values for these posterior predictive checks and they are far from 0 or 1, so this is a good fitting model.

# Variable Importance

After fitting the model with random slopes with informative priors, the importance of the covariates changed drastically in terms of what is most important. The mean, standard deviation, and 95% credible set on each parameter is displayed for both the posterior of model 1 and model 3 after pooling over the storm IDs. In the model that treated the slopes as constant across storms, every parameter had a significant effect other than MINSLP according the the 95\% credible interval. However, in model 3 when the slopes are treated as random effects and estimated from the data, the only important variables are HWRF and HWFI. We were told that HWRF would be a very important variable, but we may have another one that can improve the predictions of VMAX across storms. Since VMAX is log-normally distributed, both HWRF and HWFI increase VMAX about 8\%.

# Prediction

The posterior predictive distributions were output for the $\beta_{js}$ and $\sigma_i$ parameters. Using these PPD's to draw new values for the new values for the covariates and StormIDs that were not present in the training data set, new data was simulated. For each new observation, the posterior predictive distribution was determined and posterior predictive checks were conducted for the mean and 95\% credible set from $S = 20000$ iterations of MCMC sampling. The Bayesian p-values from the posterior predictive checks were 0.35029940, 0.68263473, and 0.04041916 for the mean, lower bound, and upper bound, respectively when compared to the observed values of the original data set. This is not a one-to-one comparison, but gives a good idea of a decent fit and prediction of the new observations.